High Performance P$^{3}$M N-body code: CUBEP$^3$M

### 2012-08-21 MNRAS_V1/paper_new.tex
Entered comments by Ue-Li

\begin{abstract}
This paper presents {\small CUBEP$^3$M}, a high performance, publicly-available, cosmological N-body code
and describes many utilities and extensions that have been added to the standard package, including a runtime halo finder,
a non-Gaussian initial conditions generator, a tuneable accuracy, and a system of unique particle identification.
{\small CUBEP$^3$M} is fast, has a low memory imprint, and 
has been run on up to $20,000$ cores, achieving close to ideal weak scaling
even at this scale.
It is well suited and has already been used for a broad number of science applications that 
require either large samples of non-linear realizations or 
very large dark matter N-body simulations,
including  cosmological reionization, baryonic acoustic oscillations, weak lensing or
non-Gaussian statistics.
We discuss the structure, the accuracy, any known systematic effects, and the scaling performance
of the code and its utilities, when applicable.
\end{abstract}

@@

 The standard package also contains support for gas cosmological evolution through a portable TVD-MHD module \citep{2003ApJS..149..447P} that scales up to thousands of cores as well (see \citet{2010arXiv1004.1680P} and footnote 4 in section \ref{sec:scaling}), and a coupling interface with the radiative transfer code C$^2$-ray \citep{2006NewA...11..374M}.
 {\small CUBEP$^3$M} is therefore one of the most competitive and versatile public N-body code, 
 and has been involved in a number of scientific applications over the last few years,
spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2012MNRAS.421..832Y, 2012arXiv1202.2332H},  BAO
 \citep{2011ApJ...728...35Z,  2012MNRAS.419.2949N, 2012MNRAS.423.2288H, 2012arXiv1205.4989H}, 
formation of early cosmic structures \citep{2008arXiv0806.2887I,2010arXiv1005.2502I},
observations of dark stars \citep{2010MNRAS.407L..74Z,2012MNRAS.422.2164I} and
reionization \citep{2012MNRAS.423.2222I,2012ApJ...750...20F,2011MNRAS.413.1353F,2012MNRAS.422..926M,2012MNRAS.424.1877D,2012MNRAS.424..762D}. 
Continuous efforts are being made to develop, extend and improve the code and each of its utilities, 
and we expect that this will pave the way to an increasing number of science projects. 
Notably, the fact that the force calculation is multi-layered makes the code extendible, 
and opens the possibility to run {\small CUBEP$^3$M} on hybrid CP2010arXiv1012.0444YU-GPU clusters.
It is thus important for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.

@@

 In terms of memory, {\small CUBEP$^3$M} can be configured such as the main footprint is dominated exclusively by the 
particle phase space array, the linked list and the buffer exchange arrays, in which case it asymptotes to about $44$ bytes per particles. 
For a fixed problem size, this can be achieved by maximizing the number of tiles, thereby reducing the size of the local fine mesh arrays.
There is a limit at which we can do such a volume breakdown, though, since the physical 
volume on each tile must be at least twice as large as the buffer, i.e. $\ge 96^3$ cells (see section \ref{sec:Poisson}).
In addition, this allows a high rate of particle exchange across nodes;  the memory per particle can be brought down to $32$ bytes 
if little particle passing is expected. 
However, minimizing the memory does not optimize the speed, as many {\small CPU}-intensive operations are performed on each tile -- {\small FFTW} for instance.
A higher number of cells per tile -- typically $176^3$ -- with fewer tiles usually run faster, in which case the other arrays in the above-mentioned list become memory heavy,
and the memory per particle is closer to 100 - 140 bytes. Also, enabling the particle ID adds about 24 bytes per particle: 8 for the main array, and 8 for the two {\small MPI}-passing buffers as well.
%uses about $130 - 150$ bytes per particles, and is in that sense more heavy than other codes like the 
For comparison, Lean-{\small GADGET-2} code uses  $84$ bytes per particles \citep{Gadget2Lean}.

@@

This paper describes {\small CUBEP$^3$M}, a public and  massively parallel P$^3$M N-body code that inherits from {\small PMFAST} 
 and that now scales well to 20,000 cores, pushing the limits of the cosmological problem size one can handle.
 This code is fast and has a memory footprint more than two times smaller than Lean-{\small GADGET-2}. 
We summarize the code structure, review the double-mesh Poisson solver algorithm, and present scaling and systematic tests
that have been performed. We also describe various utilities and extensions that come with the public release, 
including a run time halo finder, an extended pp force calculation and a non-Gaussian initial condition generator.
{\small CUBEP$^3$M} is one of the most competitive N-body code that is publicly available for cosmologists and astrophysicists,
it has already been used  for a large number of scientific applications, and it is our hope that the current documentation will 
help the community in interpreting its outcome.
The code is publicly available on github.com under {\tt cubep3m}, and extra documentation about the structure, 
compiling and running strategy is can be found on the CITA wiki page\footnote{\tt wiki.cita.utoronto.ca/mediawiki/index.php/CubePM}.

@@

The CITA simulations were run on the Sunnyvale cluster at CITA.
ITI was supported by The Southeast Physics
Network (SEPNet) and the Science and Technology Facilities Council
grants ST/F002858/1 and ST/I000976/1. 
Computations for the SciNet1024 runs were performed on the GPC supercomputer at the SciNet HPC Consortium. SciNet is funded by: the Canada Foundation for Innovation under the auspices of Compute Canada; the Government of Ontario; Ontario Research Fund - Research Excellence; and the University of Toronto. 
The authors acknowledge the
TeraGrid and the Texas Advanced Computing Center (TACC) at The
University of Texas at Austin (URL: http://www.tacc.utexas.edu) for
providing HPC and visualization resources that have contributed to the
research results reported within this paper. ITI also acknowledges the Partnership for Advanced
Computing in Europe (PRACE) grant 2010PA0442 which supported the code
scaling studies. UEP and JDE are supported by the NSERC of Canada.
VD acknowledges support by the Swiss National Science Foundation.
%{\bf ( Hugh, any financial acknowledgements to add here?)}

\bibliographystyle{mn2e}
\bibliography{mybib3_new}{}
%\bibliographystyle{amsplain}

>>>>

\begin{abstract}
This paper presents {\small CUBEP$^3$M}, a high performance, publicly-available, cosmological N-body code
and describes many utilities and extensions that have been added to the standard package, including a runtime halo finder,
a non-Gaussian initial conditions generator, a tuneable accuracy, and a system of unique particle identification.
{\small CUBEP$^3$M} is fast, has a memory imprint up to three times lower than other N-body codes, and 
has been run on up to $20,000$ cores, achieving close to ideal weak scaling
even at this scale.
It is well suited and has already been used for a broad number of science applications that 
require either large samples of non-linear realizations or 
very large dark matter N-body simulations,
including  cosmological reionization, baryonic acoustic oscillations, weak lensing or
non-Gaussian statistics.
We discuss the structure, the accuracy, any known systematic effects, and the scaling performance
of the code and its utilities, when applicable.
\end{abstract}

@@

 The standard package also contains support for gas cosmological evolution through a portable TVD-MHD module \citep{2003ApJS..149..447P} that scales up to thousands of cores as well (see \citet{2010arXiv1004.1680P} and footnote 4 in section \ref{sec:scaling}), and a coupling interface with the radiative transfer code C$^2$-ray \citep{2006NewA...11..374M}.
 {\small CUBEP$^3$M} is therefore one of the most competitive and versatile public N-body code, 
 and has been involved in a number of scientific applications over the last few years,
spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2012MNRAS.421..832Y, 2012arXiv1202.2332H},  BAO
 \citep{2011ApJ...728...35Z,  2012MNRAS.419.2949N, 2012MNRAS.423.2288H, 2012arXiv1205.4989H}, 
formation of early cosmic structures \citep{2008arXiv0806.2887I,2010arXiv1005.2502I},
observations of dark stars \citep{2010MNRAS.407L..74Z,2012MNRAS.422.2164I} and
reionization \citep{2012MNRAS.423.2222I,2012ApJ...750...20F,2011MNRAS.413.1353F,2012MNRAS.422..926M,2012MNRAS.424.1877D,2012MNRAS.424..762D}. 
Continuous efforts are being made to develop, extend and improve the code and each of its utilities, 
and we expect that this will pave the way to an increasing number of science projects. 
Notably, the fact that the force calculation is multi-layered makes the code extendible, 
and opens the possibility to run {\small CUBEP$^3$M} on hybrid CPU-GPU clusters.
It is thus important for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code.

@@

 In terms of memory, {\small CUBEP$^3$M} can be configured such as the main footprint is dominated exclusively by the 
particle phase space array, and the linked list, in which case it asymptotes to about $28$ bytes per particles. 
For a fixed problem size, this can be achieved by maximizing the number of tiles, thereby reducing the size of the local fine mesh arrays.
There is a limit at which we can do such a volume breakdown, though, since the physical 
volume on each tile must be at least twice as large as the buffer, i.e. $\ge 96^3$ cells (see section \ref{sec:Poisson}).
This allows a low rate of particle exchange across nodes;  in the default configuration, the {\tt send\_buf} and  {recv\_buf} arrays 
are much larger, in which case the memory per particle goes up to $44$ bytes. 
Enabling the particle ID adds at least 8 bytes per particle from the main array, and as the {\small MPI}-communication buffers are allocated larger sizes,
they contribute at most an additional 16 bytes as well. Hence a `lean' version with the ID enable uses down to $36$ bytes per particles.
In practice, minimizing the memory does not optimize the speed, as many {\small CPU}-intensive operations are performed on each tile -- the {\small FFTW} for instance.
A higher number of cells per tile -- typically $176^3$ -- with fewer tiles usually run faster, in which case the other arrays in the above-mentioned list become memory heavy,
and the memory per particle is closer to 100 - 140 bytes. 
%uses about $130 - 150$ bytes per particles, and is in that sense more heavy than other codes like the 
For comparison, Lean-{\small GADGET-2} code uses  $84$ bytes per particles \citep{Gadget2Lean}.

@@

This paper describes {\small CUBEP$^3$M}, a public and  massively parallel P$^3$M N-body code that inherits from {\small PMFAST} 
 and that now scales well to 20,000 cores, pushing the limits of the cosmological problem size one can handle.
 This code is fast and has a memory footprint up to three times smaller than Lean-{\small GADGET-2}. 
We summarize the code structure, review the double-mesh Poisson solver algorithm, and present scaling and systematic tests
that have been performed. We also describe various utilities and extensions that come with the public release, 
including a run time halo finder, an extended pp force calculation and a non-Gaussian initial condition generator.
{\small CUBEP$^3$M} is one of the most competitive N-body code that is publicly available for cosmologists and astrophysicists,
it has already been used  for a large number of scientific applications, and it is our hope that the current documentation will 
help the community in interpreting its outcome.
The code is publicly available on github.com under {\tt cubep3m}, and extra documentation about the structure, 
compiling and running strategy is can be found on the CITA wiki page\footnote{\tt wiki.cita.utoronto.ca/mediawiki/index.php/CubePM}.

@@

The CITA simulations were run on the Sunnyvale cluster at CITA.
ITI was supported by The Southeast Physics
Network (SEPNet) and the Science and Technology Facilities Council
grants ST/F002858/1 and ST/I000976/1. 
Computations for the SciNet1024 runs were performed on the GPC supercomputer at the SciNet HPC Consortium. SciNet is funded by: the Canada Foundation for Innovation under the auspices of Compute Canada; the Government of Ontario; Ontario Research Fund - Research Excellence; and the University of Toronto. 
The authors acknowledge the
TeraGrid and the Texas Advanced Computing Center (TACC) at The
University of Texas at Austin (URL: http://www.tacc.utexas.edu) for
providing HPC and visualization resources that have contributed to the
research results reported within this paper. ITI also acknowledges the Partnership for Advanced
Computing in Europe (PRACE) grant 2010PA0442 which supported the code
scaling studies. ULP and JDE are supported by the NSERC of Canada.
VD acknowledges support by the Swiss National Science Foundation.
%{\bf ( Hugh, any financial acknowledgements to add here?)}

%\bibliographystyle{mn2e}
\bibliographystyle{hapj}
\bibliography{mybib3_new}{}
